\documentclass{article}

\begin{document}

\title{RFC: Widget Plans}
\maketitle


\section*{Some ideas}

This document constitutes a request for comments for the Bioconductor
widget project.
The goals of this project are to provide software infrastructure and a
programming paradigm to help package developers implement widgets.
We are not interested directly in graphical user interfaces (GUIs)
since we understand that terminology to describe a fully featured
programmatic user interface. Our concern is with providing small-scale
navigable graphical interfaces for particular tasks. Examples of such
tasks include browsing (e.g. for files) or data entry. Such
tasks can be package specific in some cases while in others they can
be more generally useful.

It is our intention to implement all structure described here as S4
classes. This will provide a great deal of structure and can provide a
basis for extension by others with different needs. In the preliminary
implementation they will be lists. However, we have used the notion of
abstract data types and that should let us reuse much of the
code when the methods and classes are implemented.

While our initial implementation of some of these ideas is in
\verb+tcl/tk+ our intention is for the methodology to be general and
hence applicable to other systems such as Gtk and Java.

We begin with some definitions.

A primitive widget or {\em pWidget} is simply a set of linked interactive
tools.
Some examples:
\begin{itemize}
\item A text string, a type-in box and a button.
\item A set of linked radio buttons
\item A set of linked check boxes
\item A list-box
\item A button that can have an arbitrary R function associated with
  it. Text and a type-in box or list box may also be available.
\end{itemize}

A simple object system can be used to organize and describe each
pWidget.
Initially this will simply be a list where each element of the list
describes one component of the pWidget.
Thus, in the first pWidget suggested the list representation would be
of length three. The first component would indicate that it was a text
string and provide the text to be displayed. The second element would
be a type-in box and would contain a component describing the default
value. The third component would be button. It would have information
regarding the text to appear in the button and the function to be
evaluated if the button is pushed by the user.

%%FIXME: can pWidgets be rendered on their own or only as components
%%of other things?
A pWidget can then be easily assembled and rendered. 
All pWidgets have both a \verb+Cancel+ button and an \verb+End+
button. When the user pushes the \verb+End+ button the current values
for all components are saved in the appropriate slots and the pWidget
list is returned. Thus, the programmer has access to the users choices
and can use them to guide the evaluation of their program.

pWidgets are not themselves a complex enough structure to adequately
handle all tasks that we would like to address with this
system. Therefore we will introduce a number of more complex
components as well.
The first of these is the {\em widget}.

A {\em widget} is an ordered collection (list) of pWidgets.
To construct a widget the programmer creates and then combines a
number of pWidgets into a single list or structure. The order is
important and will determine the order in which the pWidgets will be
rendered.
A widget is contained within a single dialog box.
The default values are given by the input list and on exit the state
of the different pWidgets is recorded in the list, which is returned.

A widget can have two hooks which are not themselves rendered. 
A hook is simply a set of expressions that are evaluated at a
particular time they do not accept any arguments and hence must be
self-contained. 
Widgets can have a pre-rendering hook to handle specific set-up tasks
and a post-rendering hook to handle any tasks (such as unlinking of
temporary files) that need to be done after the widget is exited.
Any hook that is present is guaranteed to run regardless of the type
of exit.

Widgets are rendered by the \verb+widgetRender+ function.
It returns an object of the \verb+iWidget+ class which is basically a
list of lists. If the exit is normal (ie via the {\em end} button)
then the users changes will be contained in the return value and an
element {\tt end} whose value will be {\tt "END"}. 
For all other forms of exit the input
will be returned. The rendered widget by default will have a
\verb+Cancel+ button and an \verb+End+ button if there is no
definition for an end button in the control buttons list. 

Widgets themselves need to be extended to deal with more complex
interactions. A familiar prototype that seems to be generally useful
is the data entry wizard that appears when one wants to import data
into a Microsoft Excel Worksheet. The wizard is basically a set of
linked panes that guide the user through the steps of entering their
data. In order to mimic that process and other related ones we
introduce the further concept of chained widgets or {\em cWidget}s.

Basically a cWidget is simply a list of widgets that will
be rendered according to some specification. 
In the case of the data entry wizard one would like to guide the user
through a sequence of steps and that process should be enabled. In
other similar situations there will be no specific way to order the
users progression and a different sort of iteration will be needed. 
cWidgets will provide sufficient functionality to accomodate all such
uses.

The \verb+cWidgets+ will also have pre and
post rendering hooks (these are evaluated once for the entire
process). 
Some special rendering will occur depending on the type of cWidget
that is being processed. If the widgets must be processed in order
(e.g. a wizard) then only the active pane will be displayed. The
active pane will contain (as appropriate) buttons for \verb+Next+ 
and \verb+Previous+. All panes will contain a \verb+Cancel+ button and
the last widget will also have an \verb+End+ button. A counter
will be used to maintain state and for navigation. 
There will probably be options to control the display of the
\verb+Previous+ button to situations where it is appropriate. 

In this context we will also need to exercise some caution in
determining the appropriate behavior of any hooks that exist on the
widgets themselves. If the user navigates sequentially through the
widgets then their behavior is basically unambiguous but if backing up
is allowed that may require some substantially different behavior and
possible a different mechanism.

If the user pushes \verb+cancel+ the whole thing exits (again, we need
some decision about what gets returned). Here one might want to return
both the exit status and the list of widgets actually visited.
We may simply want the exit status but not the list of
widgets visited.
It is difficult to envisage a workable solution without some
experience. We hope that be implementing cWidgets very abstractly we
will be able to adopt a number of useful modifications as the project
progresses.

A paged control is a type of widget where the different
controls are represented by tabs. They can be selected in any order
and navigated as desired by the user.

Another suggestion is that there should be platform specific labels.
I think that this could be done by using \verb+options+ and installing
some package specific ones for {\em tkWidgets}.
Standard nomenclature is different depending on which platform is
being used and we would prefer to use the appropriate names for the
platform that is being used. 

\section*{Some specifics}

An example of what seems to be easy and useful is the following.
We can have a list of the form
\begin{verbatim}
list(Wtype="Typein", Name="aaa", Value="", ButtonText = "Browse",
     ButtonFun=fileBrowser) 
\end{verbatim}
This would get rendered as a typein box with name \verb+aaa+. The box
would initially be empty since \verb+Value+ was set to \verb+""+
(which is the empty string). There is a button labeled
\verb+Browse+ off the end of the typein-box that is linked to the
\verb+fileBrowser+ function. This button when clicked will evaluate
the \verb+fileBrowser+ function.
The returned value of this function will be placed in the typein-box.

It seems that an additional layer
of abstraction will be beneficial. 
To see the potential problem, suppose that the Button is linked to
\verb+objectBrowser+ and that the user has selected multiple
objects. How do we associate those with the text in the typein box?
It does not appear that we can do that directly. An extra layer of
abstraction might be enough for most purposes. A \verb+toText+
function takes any value and transforms it into a format suitable for
the text box. In the example given that transformation would be to
concatenate the multiple values with a separator (generally a comma).
There may be a need for a similar function which extracts the data
from a typein-box.

A typein-box \verb+pWidget+ has the following slots:
\begin{itemize}
\item Name: the print name for the item
\item Value: a value associated with this item
\item toText: a function that takes the value and turns it into a text
  string suitable for rendering in the text box.
\item canEdit: a boolean, can the user edit this box directly or only
  make changes via the button
\item buttonFun: a function associated with the button, the return
  value is put into Value
\item buttonText: the text to appear in the button
\item fromText: a function that takes the text from the text box and
  turns it into a value for the Value slot. If the user is allowed to
  type things in then this will be needed.
\end{itemize}

%%FIXME: Jianhua, we need to specify the contents of other pWidgets

We will need a boolean vector for the set of pWidgets to monitor
changes. The return value will be a list with components that include
\begin{itemize}
\item wList: the returned input list with values changed.
\item exit: a character string indicating which button was pushed,
  maybe even error?
\end{itemize}

When a button is clicked and a return value retrieved it is put into
the Value slot and toText is run on it to ensure the text box has up
to date data.

%%FIXME: what are frames?
At some point we need to worry about the layout of the primitive
widgets on a widget to be built by \verb+widgetRender+. tk has three
types of geometry managers namely pack, grid, and place. {\em Pack} puts
widgets in four sides namely top, right, bottom, and left. {\em Grid} puts
widgets in grids, and {\em place} places widgets at specified locations
based on absolute or relative values. For our purpose, pack and grid
would be our options. With the help of frames, we should be able to
have the widgets rendered nicely on a widget.

%%FIXME: I think that with some examples of radio buttons, list boxes
%%etc we might be able to dress this up as a submission to JSS or a
%%similar software journal. Rnews is another possibility. But I think
%%that we have done enough work for JSS.

\end{document}










